### 2015-11-02 manuscript.Rnw
fix typo/formatting

While phylogenies have been getting easier to build, it has been difficult to
re-use, combine, and synthesize the information they provide because published
trees are often only available as image files, and taxonomic information is not
standardized across studies. The Open Tree of Life (OTL) project addresses these
issues by providing a digital tree that encompasses all organisms built by
combining taxonomic information and published phylogenies. The project also
provides tools and services to query and download parts of this synthetic tree,
as well as the source data used to build it. Here, we present \texttt{rotl}, an R
package to search and download data from the Open Tree of Life directly in
R. \texttt{rotl} uses common data structures allowing researchers to take
advantage of the rich set of tools and methods that are available in R to
manipulate, analyze, and visualize phylogenies.

@@

Ideally, the ever increasing number of published phylogenies would contribute to
a synthesis of phylogenetic knowledge, ultimately leading to a better understanding of
the history of life while at the same time providing high-quality phylogenetic information 
for use in comparative analyses. However, in practice synthesizing phylogenetic data
is a difficult task. Phylogenetic information is largely scattered, often only available as image
files within publications, and the lack of standardization to store and
represent phylogenetic data makes it difficult for researchers to access,
synthesize, and integrate this information into their own research
(\citealt{Stoltzfus2012}; \citealt{Drew2013}; but see \citealt{Cranston2014} for suggestions of best
practices).

The Open Tree of Life (OTL) project aims at assembling and synthesizing our
current understanding of phylogenetic relationships across all organisms on
Earth \citep{Hinchliff2015}, while providing tools and services that facilitate
access to this information. OTL combines taxonomic information that serves as
the backbone for the phylogenetic relationships, and published phylogenies to
elucidate relationships among taxa. This combination of information is used to
structure the comprehensive synthetic tree. Studies can be contributed to the synthetic 
tree through a curator interface (\url{https://tree.opentreeoflife.org/curator}),
allowing the synthetic tree to be continuously updated as relationships are
elucidated or reevaluated. The current draft of the OTL synthetic tree contains
2.3 million tips. Beyond obvious applications across the life sciences to
explore questions in evolution, biodiversity, and conservation, the resources
OTL provides is useful for education and outreach (e.g., illustrating course
material, or developing outreach activities to explore relationships among species).

@@

The R programming language is a popular tool for  phylogenetics and comparative
analysis. The R packages \texttt{ape} \citep{Paradis2004}, \texttt{phylobase}
\citep{phylobase-0.8.0}, and \texttt{RNeXML} \citep{Boettiger2015} each provide
functions to import and manipulate trees within R and save the results in
standard data formats. Additional packages including \texttt{phytools}
\citep{Revell2012}, \texttt{geiger} \citep{Pennell2014-geiger}, and \texttt{ggtree} 
\citep{Yu-ggtree} allow users to analyze and visualize data in a 
phylogenetic context (see 
\url{https://cran.r-project.org/web/views/Phylogenetics.html} for a
comprehensive list of phylogenetics packages in R). In addition to packages for
phylogenetic and comparative analyses, a growing number of R packages allow users to query and
access data from the web (e.g. \texttt{rFISHBASE} \citep{Boettiger2012}, \texttt{rAvis} \citep{Varela2014}, and \texttt{paleoDB} \citep{Varela2015}), meaning data 
associated with taxa in a given phylogeny can be obtained directly in R. 
These packages, combined with the language's support for literate programming 
(\citealt{Knuth1984}; \citealt{Xie2014}), make R an ideal platform for reproducible research
in phylogenetics and comparative biology, as they allow a complete record of the 
steps taken in gathering, processing, and analyzing a given data set to be
produced.

Here we present \texttt{rotl}, and R package that allows users to download
phylogenetic and taxonomic data from the OTL directly in R. \texttt{rotl} takes
advantage of Application Programming Interfaces (APIs) provided by OTL to access
sub-trees from the synthetic Open Tree, as well as the published source trees that 
contribute to the synthesis. By providing direct access to high quality
phylogenetic data in R, \texttt{rotl} fills a key gap typical comparative
analysis workflows, and extends the degree to which R supports reproducible
research in phylogenetics and comparative biology.

@@

To get the relationships among a set of taxa from the Open Tree, the taxa first
need to be matched against the OTL taxonomy OTT using the TNRS. This step retrieves
the identifiers that will be used to extract the relationships for the set of
requested taxa.

@@

\texttt{rotl} can also be used to retrieve trees accompanying studies that have
been submitted through the curator interface, and identify the trees that
contribute to the synthetic tree. These trees provide a useful resource to
reproduce or expand on a previously published analysis, or to explore how the
elucidation of relationships within a clade has changed through time.

@@

The package vignettes provide two demonstrations of the integration of a
phylogeny and data associated the taxa it represents. Specifically, the 
``Data mashups'' vignette provides an example of how data associated with tip-taxa
can be gathered and visualized. Another vignette titled ``meta-analysis'' 
demonstrates how a complete analysis, including the gathering of data and a
phylogeny, can be performed in a single R session. We further demonstrate the
integration of phylogenetic data retrieved using \texttt{rotl} with other 
data here. We show how we can obtain a map of the occurrences for some of the cat species
(genus \textit{Lynx}) included in the phylogeny above using records for these
%species found in GBIF (Figure~\ref{fig:cat_map}; code available in the Appendix).

@@

\texttt{rotl} is free, open source, and released under a Simplified
BSD license. Stable versions are available from the CRAN repository
(\url{https://cran.r-project.org/package=rotl}), and development
versions are available from GitHub
(\url{https://github.com/ropensci/rotl}). The package is under active development,
and authors welcome bug reports of feature requests via the github repository.
The source for this manuscript is available on GitHub (\url{https://github.com/fmichonneau/rotl-ms}).

>>>>

While phylogenies have been getting easier to build, it has been difficult to
re-use, combine, and synthesize the information they provide because published
trees are often only available as image files, and taxonomic information is not
standardized across studies. The Open Tree of Life (OTL) project addresses these
issues by providing a digital tree that encompasses all organisms, built by
combining taxonomic information and published phylogenies. The project also
provides tools and services to query and download parts of this synthetic tree,
as well as the source data used to build it. Here, we present \texttt{rotl}, an
R package to search and download data from the Open Tree of Life directly in
R. \texttt{rotl} uses common data structures allowing researchers to take
advantage of the rich set of tools and methods that are available in R to
manipulate, analyze, and visualize phylogenies.

@@

Ideally, the ever increasing number of published phylogenies would contribute to
a synthesis of phylogenetic knowledge, ultimately leading to a better
understanding of the history of life while at the same time providing
high-quality phylogenetic information for use in comparative analyses. However,
in practice, synthesizing phylogenetic data is a difficult task. Phylogenetic
information is largely scattered, often only available as image files within
publications, and the lack of standardization to store and represent
phylogenetic data makes it difficult for researchers to access, synthesize, and
integrate this information into their own research (\citealt{Stoltzfus2012};
\citealt{Drew2013}; but see \citealt{Cranston2014} for suggestions of best
practices).

The Open Tree of Life (OTL) project aims at assembling and synthesizing our
current understanding of phylogenetic relationships across all organisms on
Earth \citep{Hinchliff2015}, while providing tools and services that facilitate
access to this information. OTL combines taxonomic information that serves as
the backbone for the phylogenetic relationships, and published phylogenies to
elucidate relationships among taxa. This combination of information is used to
structure the comprehensive synthetic tree. Studies can be contributed to the
synthetic tree through a curator interface
(\url{https://tree.opentreeoflife.org/curator}), allowing the synthetic tree to
be continuously updated as relationships are elucidated or reevaluated. The
current draft of the OTL synthetic tree contains 2.3 million tips. Beyond
obvious applications across the life sciences to explore questions in evolution,
biodiversity, and conservation, the resources OTL provides are useful for
education and outreach (e.g., illustrating course material, or developing
outreach activities to explore relationships among species).

@@

The R programming language is a popular tool for phylogenetics and comparative
analysis. The R packages \texttt{ape} \citep{Paradis2004}, \texttt{phylobase}
\citep{phylobase-0.8.0}, and \texttt{RNeXML} \citep{Boettiger2015} each provide
functions to import and manipulate trees within R and save the results in
standard data formats. Additional packages including \texttt{phytools}
\citep{Revell2012}, \texttt{geiger} \citep{Pennell2014-geiger}, and
\texttt{ggtree} \citep{Yu-ggtree}, allow users to analyze and visualize data in
a phylogenetic context (see
\url{https://cran.r-project.org/web/views/Phylogenetics.html} for a
comprehensive list of phylogenetics packages in R). In addition to packages for
phylogenetic and comparative analyses, a growing number of R packages allows
users to query and access data from the web (e.g. \texttt{rFISHBASE}
\citep{Boettiger2012}, \texttt{rAvis} \citep{Varela2014}, and \texttt{paleoDB}
\citep{Varela2015}), such that data associated with taxa in a given phylogeny
can be obtained directly in R.  These packages, combined with the language's
support for literate programming (\citealt{Knuth1984}; \citealt{Xie2014}), make
R an ideal platform for reproducible research in phylogenetics and comparative
biology, as they allow a complete record of the steps taken in gathering,
processing, and analyzing a given data set to be produced.

Here we present \texttt{rotl}, and R package that allows users to download
phylogenetic and taxonomic data from the OTL directly in R. \texttt{rotl} takes
advantage of OTL's Application Programming Interfaces (APIs) to access sub-trees
from the synthetic Open Tree, as well as the published source trees that
contribute to the synthesis. By providing direct access to high quality
phylogenetic data in R, \texttt{rotl} fills a key gap in typical comparative
analysis workflows, and extends the degree to which R supports reproducible
research in phylogenetics and comparative biology.

@@

To get the relationships among a set of taxa from the Open Tree, the taxa first
needs to be matched against the OTL taxonomy (OTT) using the TNRS. This step
retrieves the identifiers that will be used to extract the relationships for the
set of requested taxa.

@@

\texttt{rotl} can also be used to retrieve trees accompanying studies that have
been submitted through the curator interface, and identify the trees that
contribute to the synthetic tree. These trees constitute a useful resource to
reproduce or expand on a previously published analysis, or to explore how the
elucidation of relationships within a clade has changed through time.

@@

The package vignettes provide two demonstrations of the integration of a
phylogeny and data associated the taxa it represents. Specifically, the ``Data
mashups'' vignette provides an example of how data associated with tip-taxa can
be gathered and visualized. Another vignette titled ``meta-analysis''
demonstrates how a complete analysis, including the gathering of data and a
phylogeny, can be performed in a single R session. We further demonstrate the
integration of phylogenetic data retrieved using \texttt{rotl} with other data
here. We show how we can obtain a map of the occurrences for some of the cat
species (genus \textit{Lynx}) included in the phylogeny above using records for
these species found in GBIF (Figure~\ref{fig:cat_map}; code available in the
Appendix).

@@

\texttt{rotl} is free, open source, and released under a Simplified BSD
license. Stable versions are available from the CRAN repository
(\url{https://cran.r-project.org/package=rotl}), and development versions are
available from GitHub (\url{https://github.com/ropensci/rotl}). The package is
under active development, and authors welcome bug reports of feature requests
via the GitHub repository.  The source for this manuscript is available on
GitHub (\url{https://github.com/fmichonneau/rotl-ms}).
